---
title: "Spotify Wrapped: R and ggplot2 Edition"
author: "Cole Baril"
date: "`r format(Sys.Date())`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
    code_download: true
---

<!--- This style block is to tweak the formatting of the data tables in the report --->
<style>

  h1 {
  font-weight: bold;
}

  h2 {
  font-weight: bold;
  }

  table {
    display: block;
    overflow: auto;
  }

  blockquote {
    background-color: #1DB954;
    color: white;
    font-weight: bold;
 }

</style>

# Setup

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.asp = 0.8, out.width = "100%", message = FALSE, warning = FALSE)
```

## Load packages

```{r message = FALSE, warning = FALSE, echo = FALSE}

# Libraries we need
libs <- c(
  "tidyverse", "jsonlite", "knitr", "gghighlight",
  "plotly", "hms", "lubridate", "here", "gt", "webshot"
)


# install missing libraries
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == F)) {
  install.packages(libs[!installed_libs])
}

# load libraries
invisible(lapply(libs, library, character.only = T))
```

## Import & Clean

```{r}
StreamHistory0 <- fromJSON(here("Zip/MyData/endsong_0.json"), flatten = TRUE)
StreamHistory1 <- fromJSON(here("Zip/MyData/endsong_1.json"), flatten = TRUE)
StreamHistory2 <- fromJSON(here("Zip/MyData/endsong_2.json"), flatten = TRUE)
StreamHistory3 <- fromJSON(here("Zip/MyData/endsong_3.json"), flatten = TRUE)
StreamHistory4 <- fromJSON(here("Zip/MyData/endsong_4.json"), flatten = TRUE)
StreamHistory5 <- fromJSON(here("Zip/MyData/endsong_5.json"), flatten = TRUE)
StreamHistory6 <- fromJSON(here("Zip/MyData/endsong_6.json"), flatten = TRUE)

streamingData <- rbind(StreamHistory0, StreamHistory1, StreamHistory2, StreamHistory3, StreamHistory4, StreamHistory5, StreamHistory6)

spotify <- streamingData %>% 
  as_tibble() %>%
  separate(col = "ts", 
           into = c("date","time"),
           sep = "T") %>% 
  separate(col = "time",
           into = "time",
           sep = "Z") 

datetime <- as.POSIXct(paste(spotify$date, spotify$time), format = "%Y-%m-%d %H:%M:%S")
spotify$datetime <- datetime

spotify <- spotify %>% 
  mutate(datetime = datetime - hours(6)) %>% # Convert time zones to CST
  mutate(date = floor_date(datetime, "day") %>% # Creating date 
           as_date, minutes = ms_played / 60000) %>% # Convert ms played to minutes played
  mutate(time = floor_date(datetime, "minutes")) # Remove seconds from time

spotify %>% 
  glimpse(24)


# Set time limits for use in plots
lims <- as.POSIXct(strptime(c("2014-10-27", "2022-11-23"), 
                   format = "%Y-%m-%d"))

spotify$date <- ymd(spotify$date)
```


# Genre 

```{r}
spotify <- spotify %>% 
  mutate(genre = case_when(master_metadata_album_artist_name %in% c("Armin van Buuren",
                                                                    "GAIA",
                                                                    "Aly & Fila",
                                                                    "Alex M.O.R.P.H",
                                                                    "Gareth Emery",
                                                                    "Above & Beyond",
                                                                    "ReOrder",
                                                                    "Andrew Rayel",
                                                                    "Super8 & Tab",
                                                                    "Craig Connelly") ~  "Trance",
                           master_metadata_album_artist_name %in% c("Lady Gaga",
                                                                    "Halsey") ~ "Pop"))

artist_summary <- spotify %>% 
  group_by(master_metadata_album_artist_name) %>% 
  tally()
```

# Listening Patterns Over Time

## Daily 

```{r}
spotify %>% 
  filter(ms_played >= 1000) %>% 
  group_by(date)  %>%
  # group_by(date = floor_date(date, "day")) %>% # This does not work after updates. 
  summarize(songs = n()) %>% 
  arrange(date) %>% 
  ggplot(aes(x = date, y = songs)) +
  geom_col(aes(fill = songs, colour = songs)) + # Use `colour = ` here because using `fill = ` does not work with small lines.
  scale_x_date(breaks = "1 year", 
                   date_labels = "%Y",
                   expand = c(0, 0)) + # Removes white space around the plot on the x axis 
  scale_fill_gradient(high = "#0b3e34", low = "#1db954") +
  scale_colour_gradient(high = "#0b3e34", low = "#1db954") + 
  labs(x = "Date",
       y = "Number of Songs",
       colour = "Songs") +
  guides(fill = "none") +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) 
```

## Monthly

```{r}
spotify %>% 
  filter(ms_played >= 1000) %>% 
  group_by(date) %>%
  group_by(date = floor_date(date, "month")) %>% 
  summarize(songs = n()) %>% 
  arrange(date) %>% 
  ggplot(aes(x = date, y = songs)) +
  geom_col(aes(fill = songs, colour = songs)) + # Use `colour = ` here because using `fill = ` does not work with small lines.
  scale_x_date(breaks = "1 year", 
                   date_labels = "%Y",
               expand = c(0, 0)) + # Removes white space around the plot on the x axis 
  scale_fill_gradient(high = "#0b3e34", low = "#1db954") +
  scale_colour_gradient(high = "#0b3e34", low = "#1db954") + 
  labs(x = "Date",
       y = "Number of Songs",
       colour = "Songs") +
  guides(fill = "none") +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) 
```

## Listening Heat map

> This heat map shows the *total* number of minutes spent listening to Spotify during each hour in the day on each weekday. 

```{r}
spotify %>% 
  group_by(date, hour = hour(datetime), weekday = wday(date, label = TRUE)) %>% 
  summarize(minuteslistened = sum(minutes)) %>% 
  mutate(year = format(date, "%Y")) %>% 
  group_by(hour, weekday) %>% 
  summarize(minuteslistened = sum(minuteslistened)) %>% 
  drop_na() %>% # NA Weekday value of 181 dropped. 
  ggplot(aes(weekday, hour, fill = minuteslistened)) +
  geom_tile(colour = "white", size = 0.1) + 
  scale_fill_gradient(high = "#0b3e34", low = "#1db954") +
  scale_y_continuous(trans = "reverse") +
  theme_bw() +
  labs(x = "Weekday",
       y = "Hour of the Day (24-Hour, CST)",
       fill = "Minutes",
       title = "Spotify Weekly Listening Heatmap",
       subtitle = "2014 to 2022") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

```

# Artists 

> The data set is **Very** large, and `ggplot` cannot deal with very large data sets. Therefore, I will be extracting small subsets from the data. Additionally,  **Podcasts** are not considered a song with an artist, so these are excluded (for now). 

## Top 20 Artists - All Time

```{r}
spotify %>% 
  mutate(year = floor_date(date, "year")) %>% 
  mutate(year = format(year, "%Y")) %>% 
  group_by(master_metadata_album_artist_name) %>% 
  summarize(hours = sum(minutes) / 60) %>% 
  drop_na() %>% 
  arrange(desc(hours)) %>% 
  head(20, hours) %>% 
  gt() %>% 
  tab_header(title = "Top 20 Artists All Time") %>% 
  cols_label(
    master_metadata_album_artist_name = "Artist",
    hours = "Hours Listened") %>% 
  fmt_number(columns = hours,
             rows = everything(),
             decimals = 0) %>% 
  as_raw_html()
  
```

## Listening Activity by Artist Over Time

```{r}
spotify %>% 
  group_by(master_metadata_album_artist_name, date = floor_date(date, "month")) %>% 
  filter(master_metadata_album_artist_name %in% c("Armin van Buuren", 
                                                  "GAIA",
                                                  "Above & Beyond",
                                                  "Gareth Emery",
                                                  "Alex M.O.R.P.H.",
                                                  "Gustav Mahler",
                                                  "Lady Gaga",
                                                  "ReOrder",
                                                  "Halsey",
                                                  "Aly & Fila")) %>% 
  summarize(hours = sum(minutes) / 60) %>% 
  drop_na() %>% 
  ggplot(aes(x = date, y = hours, group = master_metadata_album_artist_name, 
             colour = master_metadata_album_artist_name)) +
  geom_line() +
  scale_x_date(breaks = "1 year", date_labels = "%Y") +
  gghighlight(master_metadata_album_artist_name == "Armin van Buuren" || 
              master_metadata_album_artist_name == "Gustav Mahler" ||
              master_metadata_album_artist_name == "Lady Gaga" ||
              master_metadata_album_artist_name == "Halsey") +
  labs(title = "Hours Listened to Most Listened Artists Over Time",
       subtitle = "Data for top 10 artists only \nData aggregated by month. Other grey lines are top Trance artists",
       x = "Date", 
       y = "Hours Listened") +
  theme_bw()  +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

# Top 20 Songs

```{r}
spotify %>% 
  mutate(year = floor_date(date, "year")) %>% 
  mutate(year = format(year, "%Y")) %>% 
  group_by(master_metadata_track_name) %>% 
  summarize(hours = sum(minutes) / 60) %>% 
  drop_na() %>% 
  arrange(desc(hours)) %>% 
  head(20, hours) %>% 
  gt() %>% 
  tab_header(title = "Top 20 Songs All Time") %>% 
  cols_label(
    master_metadata_track_name = "Song",
    hours = "Hours Listened") %>% 
  fmt_number(columns = hours,
             rows = everything(),
             decimals = 0) 
```

## Most Listened to Artist in a Day

```{r}
spotify %>% 
  group_by(master_metadata_album_artist_name, 
           date = floor_date(date, "day")) %>% 
  summarize(hours = sum(minutes) / 60) %>% 
  arrange(desc(hours)) %>% 
  drop_na() %>% 
  head(100, hours) %>% 
  gt()

```

# Listening Device 

```{r}


device_time <- spotify %>% 
  mutate(device = case_when(str_detect(platform, 'Android') ~ 'Android',
                            str_detect(platform, 'iOS') ~ 'iPhone',
                            str_detect(platform, 'Windows|windows|web') ~ 'Windows',
                            str_detect(platform, 'google') ~ 'Google Home'
                            )) %>% 
  group_by(device) %>% 
  summarise_at(vars(minutes), sum) %>% 
  mutate(hours = minutes/60) %>% 
  mutate(days = hours/24)

device_time %>% 
  add_row(
    device_time %>% 
    summarise(across(2:last_col(), sum) %>% 
    mutate(device = "Total"))
    ) %>% 
  gt() %>% 
  tab_header(title = "Time Spent Listening by Platform") %>% 
  cols_label(
    device = "Platform",
    minutes = "Minutes",
    hours = "Hours",
    days = "Days") %>% 
  fmt_number(columns = c(minutes, hours, days),
             rows = everything(),
             decimals = 0) %>% 
  as_raw_html()

date_minutes <- spotify %>% 
  mutate(device = case_when(str_detect(platform, 'Android') ~ 'Android',
                            str_detect(platform, 'iOS') ~ 'iPhone',
                            str_detect(platform, 'Windows|windows|web') ~ 'Windows',
                            str_detect(platform, 'google') ~ 'Google Home'
                            )) %>% 
  group_by(device, date) %>% 
  summarise_at(vars(minutes), sum) %>% 
  mutate(hours = minutes/60) %>% 
  mutate(days = hours/24)

```

# References

```{r}
library(purrr)
c("tidyverse", "jsonlite", "knitr", "gghighlight",
  "hms", "lubridate", "here", "gt", "webshot") %>%
  map(citation) %>%
  print(style = "text")
```

